D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-29 01intro_preliminary.tex
.

Dirac notation, also known as bra-ket notation, provides an intuitive and concise mathematical framework for describing quantum states and operations in quantum mechanics. 
We write $\cH$ for a Hilbert space, i.e., a vector space equipped with an standard inner produce $\<\bu,\bv\> \in \bC$ for $\bu,\bv\in\cH$.
Dirac notation consists of the following components that reflect the basic postulates of quantum mechanics:
\begin{itemize}
  \item Ket $|u\>$ is a column vector that denotes a quantum state $\bu$ in the state Hilbert space $\cH$. For example, the computational bases of qubit system are commonly written as $|0\> = \matrix{1 \\ 0}$ and $|1\> = \matrix{0 \\ 1}$.
  \item Bra $\<u|$ is a row vector, the conjugate transpose of $|u\>$, that denotes the dual state of $|u\>$. 
  %It is alternative to interpret as a linear mapping from $\cH$ to $\bC$ defined as $\<u| : |v\> \mapsto \<\bu,\bv\>$. It is the conjugate transpose of $|u\>$.
  For exmaple, $\<0| = \matrix{1 & 0}$ and $\<1| = \matrix{0 & 1}$.
  \item Inner produce $\<u|v\> \triangleq \<\bu,\bv\>$ which indicates the probability amplitude for $|u\>$ to collapse into $|v\>$. 
  In convention, it is computed by matrix multiplication of two state, e.g., 
  $\<0|1\> = \matrix{1 & 0} \matrix{0 \\ 1} = 0.$
  \item Outer product $|u\>\<v| \triangleq |w\> \mapsto (\<v|w\>) |u\>$. Any linear map, such as unitary tranformation, measurement operators, etc, can be decomposed as the sum of outer products. It is also computed by matrix multiplication, e.g., 
  $|1\>\<0| = \matrix{0 \\ 1} \matrix{1 & 0}  = \matrix{0 & 0 \\ 1 & 0}.$
  \item Tensor product $|u\>\otimes|v\>$ (or simply $|u\>|v\>$ or $|uv\>$), $\<u|\otimes\<v|$ (or simply $\<u|\<v|$ or $\<uv|$) for describing the state, dual state and linear map of composite systems respectively. It is computed by the the Kronecker product of matrices, e.g., $\<0|\<1| = {\color{blue}\matrix{1 & 0}}\otimes {\color{red}\matrix{1 & 0}} = \matrix{({\color{blue}1}*{\color{red}1}) & ({\color{blue}1}*{\color{red}0}) & ({\color{blue}0}*{\color{red}1}) & ({\color{blue}0}*{\color{red}0})} = \matrix{1&0&0&0}$.
\end{itemize}

@@

Unfortunately, the standard Dirac notation becomes unexpectedly tedious for describing multi-body systems. To see this, consider the following example:
\begin{example}
  \label{example1}
  Let $p,q,r$ be three qubits and initially in the (unnormalized) $\ghz$ state $|\ghz\> \triangleq |000\>+|111\>$. Show that applying 3-qubit Toffoli gate ($\ccnot$) with control qubits $p,r$ and target qubit $q$ is equivalent to applying 2-qubit $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$. 
\end{example}
We formalize the conclusion using standard Dirac notation as
\begin{align*}
  \ccnot|\ghz\> &= (I\otimes \cnot)|\ghz\> \\
  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\>
\end{align*}
following the steps: 1. Arrange qubits in the conventional order, here we choose $p,r,q$ to simplify the representation of $\ccnot$; 2. Lift the local operation $\cnot$ to the global system. For $\cnot$ acting on $r,q$ is easier, since $r,q$ consistent with the chosen order, and we only need to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes \cnot)$.
For $\cnot$ acting on $p,q$, realize that $p,q$ is not adjacent in the chosen order, we additionally need to use $\swap$ to temporarily exchange the qubits $p$ and $r$, i.e., globally, the $(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of $\cnot$ on $r,q$).

@@

To address this, physicists introduce labels (or subscripts) to indicate the systems on which quantum states or operations are applied, thereby avoiding unnecessary lifting and swap gates. For example, rewriting the above example using labels:
\begin{align*}
  \ccnot_{prq}|\ghz\>_{pqr} = 
  \cnot_{rq}|\ghz\>_{pqr}
  = \cnot_{pq}|\ghz\>_{pqr}
\end{align*}
without determining the conventional order nor lifting using additional $I$ and $\swap$s. 
The use of labels allows us to avoid deliberately maintaining the order of qubits. Tensor product become an AC symbol, so we can rearrange qubits as needed to perform calculations. For example, we can perform the calculation as follows:
\begin{align*}
  &\ccnot_{prq}|\ghz\>_{pqr}
  = (\ccnot|000\> + \ccnot|111\>)_{prq}
  = (|000\> + |110\>)_{prq}\\
  &\cnot_{rq}|\ghz\>_{pqr}
  = (\cnot|00\>)_{rq}|0\>_p + (\cnot|11\>)_{rq}|1\>_p
  =(|000\> + |101\>)_{rqp}\\
  &\cnot_{pq}|\ghz\>_{pqr}
  = (\cnot|00\>)_{pq}|0\>_r + (\cnot|11\>)_{pq}|1\>_r
  = (|000\> + |101\>)_{pqr}
\end{align*}
where the RHS of each line are equivalent.
In addition, it can very freely describe local measurements, partial traces (the state or evolution of subsystems in multi-body systems), and partial inner products (the representation of partial traces in pure states), which are sufficient to handle the mathematical formulas of quantum mechanics in multi-body systems.

@@

We generalize Example \ref{example1} as the motivating example in the following:
\begin{example}[Apply multiplexer to $\ghz$ state]
  \label{ex: motivating}
  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\>\<iii|$. Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$. Show that:
  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
\end{example}

>>>>

Dirac notation, also known as bra-ket notation, provides an intuitive and concise mathematical framework for describing quantum states and operations in quantum mechanics. 
We write $\cH$ for a Hilbert space, i.e., a vector space equipped with an standard inner produce $\langle\bu,\bv\rangle \in \bC$ for $\bu,\bv\in\cH$.
Dirac notation consists of the following components that reflect the basic postulates of quantum mechanics:
\begin{itemize}
  \item Ket $|u\rangle$ is a column vector that denotes a quantum state $\bu$ in the state Hilbert space $\cH$. For example, the computational bases of qubit system are commonly written as $|0\rangle = \matrix{1 \\ 0}$ and $|1\rangle = \matrix{0 \\ 1}$.
  \item Bra $\langle u|$ is a row vector, the conjugate transpose of $|u\rangle$, that denotes the dual state of $|u\rangle$. 
  %It is alternative to interpret as a linear mapping from $\cH$ to $\bC$ defined as $\langle u| : |v\rangle \mapsto \langle \bu,\bv\rangle$. It is the conjugate transpose of $|u\rangle$.
  For exmaple, $\langle 0| = \matrix{1 & 0}$ and $\langle 1| = \matrix{0 & 1}$.
  \item Inner produce $\langle u|v\rangle \triangleq \langle \bu,\bv\rangle$ which indicates the probability amplitude for $|u\rangle$ to collapse into $|v\rangle$. 
  In convention, it is computed by matrix multiplication of two state, e.g., 
  $\langle 0|1\rangle = \matrix{1 & 0} \matrix{0 \\ 1} = 0.$
  \item Outer product $|u\rangle\langle v| \triangleq |w\rangle \mapsto (\langle v|w\rangle) |u\rangle$. Any linear map, such as unitary tranformation, measurement operators, etc, can be decomposed as the sum of outer products. It is also computed by matrix multiplication, e.g., 
  $|1\rangle\langle 0| = \matrix{0 \\ 1} \matrix{1 & 0}  = \matrix{0 & 0 \\ 1 & 0}.$
  \item Tensor product $|u\rangle\otimes|v\rangle$ (or simply $|u\rangle|v\rangle$ or $|uv\rangle$), $\langle u|\otimes\langle v|$ (or simply $\langle u|\langle v|$ or $\langle uv|$) for describing the state, dual state and linear map of composite systems respectively. It is computed by the the Kronecker product of matrices, e.g., $\langle 0|\langle 1| = {\color{blue}\matrix{1 & 0}}\otimes {\color{red}\matrix{1 & 0}} = \matrix{({\color{blue}1}*{\color{red}1}) & ({\color{blue}1}*{\color{red}0}) & ({\color{blue}0}*{\color{red}1}) & ({\color{blue}0}*{\color{red}0})} = \matrix{1&0&0&0}$.
\end{itemize}

@@

Unfortunately, the standard Dirac notation becomes unexpectedly tedious for describing multi-body systems. To see this, consider the following example:
\begin{example}
  \label{example1}
  Let $p,q,r$ be three qubits and initially in the (unnormalized) $\ghz$ state $|\ghz\rangle \triangleq |000\rangle+|111\rangle$. Show that applying 3-qubit Toffoli gate ($\ccnot$) with control qubits $p,r$ and target qubit $q$ is equivalent to applying 2-qubit $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$. 
\end{example}
We formalize the conclusion using standard Dirac notation as
\begin{align*}
  \ccnot|\ghz\rangle &= (I\otimes \cnot)|\ghz\rangle \\
  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\rangle
\end{align*}
following the steps: 1. Arrange qubits in the conventional order, here we choose $p,r,q$ to simplify the representation of $\ccnot$; 2. Lift the local operation $\cnot$ to the global system. For $\cnot$ acting on $r,q$ is easier, since $r,q$ consistent with the chosen order, and we only need to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes \cnot)$.
For $\cnot$ acting on $p,q$, realize that $p,q$ is not adjacent in the chosen order, we additionally need to use $\swap$ to temporarily exchange the qubits $p$ and $r$, i.e., globally, the $(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of $\cnot$ on $r,q$).

@@

To address this, physicists introduce labels (or subscripts) to indicate the systems on which quantum states or operations are applied, thereby avoiding unnecessary lifting and swap gates. For example, rewriting the above example using labels:
\begin{align*}
  \ccnot_{prq}|\ghz\rangle_{pqr} = 
  \cnot_{rq}|\ghz\rangle_{pqr}
  = \cnot_{pq}|\ghz\rangle_{pqr}
\end{align*}
without determining the conventional order nor lifting using additional $I$ and $\swap$s. 
The use of labels allows us to avoid deliberately maintaining the order of qubits. Tensor product become an AC symbol, so we can rearrange qubits as needed to perform calculations. For example, we can perform the calculation as follows:
\begin{align*}
  &\ccnot_{prq}|\ghz\rangle_{pqr}
  = (\ccnot|000\rangle + \ccnot|111\rangle)_{prq}
  = (|000\rangle + |110\rangle)_{prq}\\
  &\cnot_{rq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{rq}|0\rangle_p + (\cnot|11\rangle)_{rq}|1\rangle_p
  =(|000\rangle + |101\rangle)_{rqp}\\
  &\cnot_{pq}|\ghz\rangle_{pqr}
  = (\cnot|00\rangle)_{pq}|0\rangle_r + (\cnot|11\rangle)_{pq}|1\rangle_r
  = (|000\rangle + |101\rangle)_{pqr}
\end{align*}
where the RHS of each line are equivalent.
In addition, it can very freely describe local measurements, partial traces (the state or evolution of subsystems in multi-body systems), and partial inner products (the representation of partial traces in pure states), which are sufficient to handle the mathematical formulas of quantum mechanics in multi-body systems.

@@

We generalize Example \ref{example1} as the motivating example in the following:
\begin{example}[Apply multiplexer to $\ghz$ state]
  \label{ex: motivating}
  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\rangle\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\rangle\langle iii|$. Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\rangle\langle ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\rangle\langle i|\otimes U_{ii}$. Show that:
  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\rangle_{pqr} = N_{pq}|\ghz\rangle_{pqr}.$$
\end{example}
